
 

# R code collection {#s07-Code-Collection} 
 
```{r, echo=F}
save(AGMeans, Amat.array, Create.Amat, DecomposeF, DecomposeU, elasticity.matrix, EstimateSGR, EulerLotka, GenTime, LC, LifeExpectancy, nsim.stochastic, nsimLC, projection, projection.stochastic, R0.lifetable, R0function, sensitivity.matrix, TC.lifetable, uvlambda, lx, mx, px, Abird.post, Abird.pre, LifeTableBird, Fbird.pre,  Ubird.pre, Pop.bird.pre, Pop.bird.post, Pop.bird.transpose, stablestructure.pre, repvalues.pre, file="CompendiumCode.RData")
```

This appendix includes a collection of the R functions used in the compendium. 

An RData-file containing all the functions as well as relevant R objects for the songbird example can be downloaded here:

```{r , echo=FALSE}
# a single file
xfun::embed_file('CompendiumCode.RData')
```

<br>

To import this file into R, use `load("CompendiumCode.RData")` (after saving the file in your R working directory folder). Then all functions and objects should be visible in the 'Environment' tab.

 
<br>

## Life tables

### Net reproductive rate $R_0$

```{r}
R0.lifetable <- function(lx, mx){
  sum(lx*mx) 
}
```


<br>

### Euler-Lotka 

Part of the method to solve the equation for $\lambda$ using `uniroot()` (see main text).

```{r}
EulerLotka <- function(lambda, x, lx, mx){
  sum(lambda^(-x)*lx*mx)-1
}
```



<br><br>

## Matrix population models

<br>

### Create Leslie matrix 

Build Leslie matrix from vectors of survival and fertility coefficients (need to have same length). If the last survival coefficient is not zero, individuals can accumulate in the final class.

```{r}
Create.Amat <- function(Svec, Fvec){
  k <- length(Svec)
  MatA<- matrix(0,k,k)
  MatA[1,] <- Fvec 
  for(i in 1:(k-1)){
    MatA[i+1,i] <- Svec[i]
  }
  MatA[k,k] <- Svec[k] 
  MatA
  }
```
 
<br>
 
 
### Decompose $\mathbf{U}$ and $\mathbf{F}$

Decompose the matrices $\mathbf{U}$ and $\mathbf{F}$ based on equation to obtain the transition matrix $\mathbf{G}$ vector of survival probability $\mathbf{s}$ (from  $\mathbf{U}$), and the transition matrix $\mathbf{Q}$  the  fertility vector $\mathbf{f}$ (from  $\mathbf{F}$)


```{r}
DecomposeU <- function(MatU){
  k<-dim(MatU)[1]
  if(is.na(MatU[k,k])){
    MatU[k,k]<-0
  }
  Svec <- apply(MatU,2,sum)
  MatG <- t(t(MatU)/Svec)
  for(i in 1:k){
    if(Svec[i]==0){
      MatG[,i] <- 0
      MatG[i,i] <- 1
    }
  }
 list("Gmat"=MatG, "Survival"=Svec)
}
```

<br>

```{r}
DecomposeF <- function(MatF){
	k <- dim(MatF)[1]
	Fvec <- apply(MatF,2,sum)
	MatQ <- matrix(0,k,k)
	for(i in 1:k){
		if(Fvec[i]==0){
			MatQ[1,i]<-1
			
		}
		else{
			MatQ[,i]<-MatF[,i]/Fvec[i]
		}
	}
 list("Qmat"=MatQ, "Fertility"=Fvec)
}
```


<br>
 
 
 
### Population projection 

Project size of stages over time given a projection matrix and an initial population vector. Return as data frame.

```{r}
projection <- function(MatA, Tmax = 50, n0){
  k <- dim(MatA)[1]
  if(length(n0)!=k){
	  warning("n0 should have length 
	          k corresponding to 
	          number of stages in Amat")
	  }
  Nmat <- matrix(NA, nrow= Tmax+1, ncol=k)
  cnames <- paste("n", 1:k, sep="")
  timesteps <- 0:Tmax
  Nmat[1,] = n0 
  for(i in 2:(Tmax+1)){  
    Nmat[i,] = MatA%*%Nmat[i-1,]
  }
  frame <- data.frame(timesteps, Nmat)
  colnames(frame) <- c("time", cnames)
  frame
}
```

 <br>
 
### Long-term growth rate, stable structure and reproductive value

For a projection matrix $\mathbf{A}$, the function returns the asymptotic growth rate $\lambda$, a vector describing the stable stage structure $\mathbf{u}$, and a vector describing the stage-specific reproductive values $\mathbf{v}'$. NB: We scale the reproductive values so that $\mathbf{v}'\mathbf{u}=1$.   

```{r}
uvlambda <- function(MatA){
 ev <- eigen(MatA)
 tev <- eigen(t(MatA))
 lmax <- which.max(Re(ev$values))
 U <- ev$vectors
 V <- tev$vectors
 u <- as.matrix(abs(Re(U[, lmax]))/
                  sum(abs(Re(U[, lmax]))))
 u <- u/(sum(u)) #scale u
 v <- as.matrix(abs(Re(V[, lmax])))
 v <- v/sum(u*v ) #scale v
 v <- t(ifelse(u*v <= 0, 0, v))
 return(list("lambda"=max(Re(ev$values)),
             "u"=u,
             "v"=v))
}
```

<br>

### Life expectancy

For a projection matrix MatU, the function returns the vector of life expectancies for given stages. Life expectancy at the first stage is the first element of this vector.

```{r}
 LifeExpectancy <- function(MatU){   
  k<-dim(MatU)[1]
  if(is.na(MatU[k,k])){
    MatU[k,k]<-0
  }
  uDim=dim(MatU)[1]
  N = solve(diag(uDim[1])-MatU)   
  colSums(N)  
}
```
<br>


### Lifetime reproduction

The function returns the mean lifetime reproduction $R_0$, based on the matrices $\mathbf{A}$ and $\mathbf{U}$. and the vector of expected remaining lifetime reproduction by stage. If all offspring are born into the first stage, the first element also corresponds to $R_0$.

```{r}
R0function  <-  function(MatU, MatF){
  k <- dim(MatU)[[1]]
  Rmat <- MatF%*%solve(diag(1, k, k)-MatU)    
  Rvec <- apply(Rmat,2,sum)
  R0 <- uvlambda(MatA=Rmat)$lam
  return(list("R0"= R0, "Rvec"= Rvec))
}
```

<br>

### Generation time

For a projection matrix MatA and a reproduction matrix MatF, the function returns the generation time estimated as the mean age of mothers in a population with stable structure.

```{r}
GenTime  <-  function(MatA, FMat){
  res <- uvlambda(MatA=MatA)
  lam <- res$lam
  u <- res$u
  v <- res$v
  lam/(v%*%FMat%*%u)    
}
```

<br>


### Sensitivities and elasticities

For a given projection matrix $\mathbf{A}$, the function `sensitivity.matrix()` returns the matrix of sensitivities  of $\lambda$ to each corresponding element of MatA, and the function  `elasticity.matrix()` returns the elasticities. The argument 'zeroes' can be set to FALSE to return also the sensitivities / elasticities corresponding to elements of MatA that are zero.


 

```{r}
sensitivity.matrix <- function(MatA, zeroes=T){
  res <- uvlambda(MatA=MatA)
  sensmat <- t(res$u%*%res$v)
  if (zeroes==T){
    sensmat <- ifelse (MatA==0, 0, sensmat)
  }
 sensmat
}
```


<br>

```{r}
elasticity.matrix <- function(MatA, zeroes=T){
  res <- uvlambda(MatA=MatA)
  sensmat <- t(res$u%*%res$v)
  if (zeroes==T){
    sensmat <- ifelse (MatA==0, 0, sensmat)
  }
 sensmat*MatA/res$lam 
}
```

<br>

 
 
## Stochastic models
 
### Arithmetic and geometric mean

The function calculates the arithmetic and geometric mean of a sequence of annual population growth rates $\Lambda_t$ with a given mean and standard deviation. It returns a data frame with time steps, the generated random sequence of annual growth rates $\Lambda_t$, and the two means.

```{r}
AGMeans <- function(Tmax=100, mean.lambda=1.05, sd.lambda=.3){
  #Use lognormal distribution to avoid negative values
  Lambdavec <- rlnorm(n=Tmax, 
                      log(mean.lambda)-
                        .5*log(1+ sd.lambda^2/
                                 (mean.lambda^2)), 
                      sdlog=sqrt(log(1+sd.lambda^2/
                                       (mean.lambda^2))))
   geomLambda <- exp(mean(log(Lambdavec)))
   meanLambda <- mean(Lambdavec)
  data.frame("Time"=0:(length(Lambdavec)-1), 
             "Lambda"=Lambdavec,
             "Geometric"=geomLambda, 
             "Arithmetic"=meanLambda)
}
```
 
 <br>
 
### Simulations in Lewontin-Cohen model
 
Function to generate one realization of the model, given initial population size, mean and standard deviation of the annual growth rates $\Lambda_t$:


```{r }
LC <- function(Tmax=30, 
               n0=100, 
               mean.lambda=1.05, 
               sd.lambda=0.1){
  stochastic.lambdas <- rnorm(Tmax, 
                              mean=mean.lambda, 
                              sd=sd.lambda)
  Nvec <- rep(NA, Tmax+1)
  Nvec[1] <- n0
  for(i in 1:Tmax){
    Nvec[i+1] <- stochastic.lambdas[i]*Nvec[i]
  }
  data.frame("Time"=0:Tmax, "N"=Nvec)
}
```


<br>


Function to generate a chosen number (`nsim`) of realizations of the Lewontin-Cohen model, returned in a data frame.


```{r}
nsimLC <- function(nsim=100, 
                   Tmax=30, 
                   n0=100, 
                   mean.lambda=1.05, 
                   sd.lambda=0.1){
  frame <-  data.frame("Time"=0:Tmax)
  for(j in 1:nsim){
  stochastic.lambdas <- rnorm(Tmax, 
                              mean=mean.lambda, 
                              sd=sd.lambda)
  Nvec <- rep(NA, Tmax+1)
  Nvec[1] <- n0
  for(i in 1:Tmax){
    Nvec[i+1] <- stochastic.lambdas[i]*Nvec[i]
  }
  frame <- cbind(frame, Nvec)
  names(frame)[j+1] <- paste("N",j,sep="")
  }
  frame
}
```


### Simulations of stochastic structured model

#### Generate sequence of projection matrices

The function below returns a time series of stochastic projection matrices as an array, for a model assuming stochasticity in survival coefficients and fertility coefficients only, and no correlation between matrix elements (can be modified to include such correlation). Fertility is modeled with a log-link (normally distributed on log-scale), while survival is modeled with a log-log-link (normally distribued on log-log-scale). The mean (constant) environment is determined by the input matrices $\mathbf{A}$, $\mathbf{U}$ and $\mathbf{F}$. The input vectors `VarF` and `VarS` specify the variances of fertility and survival on their respective scales.


```{r }
Amat.array <- function(MatA, 
                       MatU, 
                       MatF, 
                       tmax=30, 
                       VarF=rep(0.03, 3), 
                       VarS=rep(0.03, 3)){
  #Split survival/transition matrix:
  SplitU <- DecomposeU(MatU) 
  #Split fertility matrix:
  SplitF <- DecomposeF(MatF) 
   #Transition matrix:
  Gmat <- SplitU$Gmat
   #Survival vector:
  Svec <- SplitU$Survival
  #Define beta (log-log link function):
  Beta.S <- -log(-log(Svec))
  #Offspring transition matrix:
  Qmat <- SplitF$Qmat 
  #Fertility vector:
  Fvec <- SplitF$Fertility 
  #Define beta (log link function):
  Beta.F <- log(Fvec) 
 
  #variance covariance matrix 
  SigMat <- diag(c(VarS, VarF))
  #Draw random (scaled) values for S and F:
  SFTime <- mvrnorm(tmax, mu=c(Beta.S, Beta.F), Sigma=SigMat)
  #Number of matrix classes 
  k <- length(Svec) 
  #Rescaled survival probabilities (loglog link):
  SvecTime<-   exp(-exp(-SFTime[,1:k])) 
  #Rescaled fertilities  (log link):
  FvecTime<- exp(SFTime[,(k+1):(2*k)]) 
#Build  projection matrix for each time step:
  A.array <- array(NA,dim=c(k,k,tmax))
  for (i in 1:tmax){
    Smat <- diag(SvecTime[i,])
    Fmat <- diag(FvecTime[i,])
    A.array[,,i] <-   Gmat%*%Smat+Qmat%*%Fmat
  }
return("Amats" = A.array)
}
```

<br>


#### Stochastic population projection

The function below returns a data frame with the size of each stage over time, for a given array of projection matrices (third dimension should correspond to time) and initial population vector `n0`.

```{r }
projection.stochastic <-function(Amats, n0){
	Tmax <- dim(Amats)[3]
	k <- dim(Amats)[1]
	if(length(n0)!=k){
	  warning("n0 should have length k 
	          corresponding to number of 
	          stages in the Amats object")
	  }
	Nmat <- matrix(NA,nrow= Tmax+1, ncol=k)
	Nmat[1,] <- n0
  cnames <- paste("n", 1:k, sep="")
  timesteps <- 0:Tmax
	for(i in 2:(Tmax+1)){  
		Nmat[i,] <- Amats[,,i-1] %*% Nmat[i-1,]
		}
  frame <- data.frame(timesteps,Nmat)
  colnames(frame) <- c("time", cnames)
  frame
}
```


<br>


The following function generates a series of stochastic matrices `nsim` times, each series of length `Tmax`, and returns the projected total population size over time for each series, in a data frame.  Other input variables are the constant matrices and variance vectors,  and initial population vector.

```{r }
nsim.stochastic <- function(nsim=100, 
                            tmax=30, 
                            n0=rep(10,3), 
                            MatA, 
                            MatU, 
                            MatF, 
                            VarF, 
                            VarS){
  frame <-  data.frame("Time"=0:tmax)
  for(m in 1:nsim){
  AMATS <- Amat.array(MatA=MatA, 
                      MatU=MatU, 
                      MatF=MatF, 
                      VarF=VarF, 
                      VarS=VarS, 
                      tmax=tmax)
   projectN <- apply(projection.stochastic(Amats=AMATS, 
                                           n0=n0)[,-1],1,sum)
  frame <- cbind(frame, projectN)
  names(frame)[m+1] <- paste("Rep",m,sep="")
  }
  frame
}
```

<br>

### Estimate stochastic growth rate

The following code can be used to calculate the stochastic growth rate following the definition of equation \@ref(eq:StochasticRate), given one sequence of projection matrices. The cutoff argument removes the first time steps (corresponding to cutoff number).

```{r}
 
EstimateSGR <- function(Amats, n0, cutoff=20){
  Tmax <- dim(Amats)[3]
	if(cutoff>=Tmax){
	  warning("cutoff must be lower 
	          than the number of 
	          matrices in Amats")
	  }
  project <- projection.stochastic(Amats=Amats, n0=n0) 
  Ntot <- apply(project[,-1],1,sum)
  logN <- log(Ntot[cutoff:length(Ntot)])
  TM <- length(logN)
  (logN[TM]-logN[1])/TM
}

```



<br>


## Parameters and matrices for the songbird example

Copy paste if you want to look at example code without running all the previous code.
 
```{r}
#------------------------------
#Life table parameters
#------------------------------
x <- 0:3
mx <- c(0, 0, 3, 6) 
lx <- c(1, 0.2, 0.18, 0.108)
px <- c(.2, .9, .6, 0)

#------------------------------
#Life table calculations
#------------------------------
TC_Bird <- TC.lifetable(x,  lx, mx)
R0_Bird <- R0.lifetable(lx, mx) 

#------------------------------
#Apply uniroot to find lambda 
#(EulerLotka defined among funcions)
#------------------------------
lambdabird <- uniroot(EulerLotka, 
                      x=x,
                      lx=lx,
                      mx=mx,
                      interval=c(.1,2))$root

#------------------------------
#Projection matrices
#------------------------------
Abird.pre <- Create.Amat(
  Svec <- c(.9, .6, 0), 
  Fvec= c(0, .6, 1.2))  
Abird.post <- Create.Amat(
  Svec <- c(.2, .9, 0), 
  Fvec= c(0, 2.7, 3.6))  

#------------------------------
#Projection 
#------------------------------
Pop.bird.pre <- projection(
  MatA=Abird.pre,
  n0=rep(10,3), 
  Tmax=30)

Pop.bird.post <- projection(
  MatA=Abird.post, 
  n0=rep(10,3),
  Tmax=30)

```